Parameter Inference for a Simple Stochastic Hydrological Model

### 2015-03-13 TeXFiles/stochInf1D.tex
q variables from the very beginning

In this paper, we study the feasibility of a full Bayesian inference, for simple conceptual rainfall-runoff models with scale-invariant noise.
We start with a single linear reservoir and describe the dynamics of its water content, $S(t)$, by means of the stochastic differential equation
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ describes the rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the retention time and the noise.
The scale-invariance of the noise term renders the noise parameter dimensionless.
Due to the state-dependence of the noise term, eq. (\ref{sde}) is ill-defined unless we specify the discretization convention.
For computational reasons, we work with the {\em Stratonovich} convention.
Our parametrization is such that, for constant rain input $r(t)=r$ and in the long-time limit, the mean of $S$ converges to the limit of the unperturbed ($\gamma=0$) system $Kr$ (see below).

According to \cite{lau_2007_StateDepDiff}, the probability, $P(S,T|S_0,t_0)$, of finding the system in a state $S$ at time $t = T$ given that it was in an initial state $S_0$ at time $t = 0$, can be expressed in the form of a path integral as
\begin{equation}\label{pathint}
P(S,T|S_0,t_0) = \int_{S_0}^{S} e^{-\tilde{\mathcal S}[\dot{S}(t),S(t)]} \mathcal{D}S \,,
\end{equation}
where the integral extends over all paths starting at $S_0$ and ending at $S$.
The path-measure $\mathcal DS$ is the formal limit
\begin{equation}\label{pathmeasure}
{\mathcal D}S
=
\lim_{\Delta t\rightarrow 0}
\prod_{i}
\frac{dS_i}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S_i}\,,
\end{equation}
for a discretization $S_i = S(t_i)$ with $t_i-t_{i-1}=\Delta t$.
The {\em action} is a functional defined on paths $S:[0,T]\longrightarrow \mathbb R_+$ and reads as
\begin{equation}\label{action}
  \tilde{\mathcal S}[\dot{S}(t),S(t)]
  =
  \int_0^T
  dt
  \left\{
    \frac{K}{2\gamma S^2(t)}
    \left(
        \dot S(t)
        -
        r(t)
        +
        \frac{1+\gamma}{K}S(t)
    \right)^2
    -
    \frac{2+\gamma}{4K}
  \right\}
 \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $S(t)$ and ensures the normalization condition $\int{ P(S,T|S_0,t_0) \mathit{dS}}=1$.

It is possible to calculate the equilibrium distribution $P_{\text{eq}}(S) = \lim_{T\rightarrow\infty} P(S,T|S_0,t_0)$ in the simple case of a constant rain input $r_{0}$. Systems at thermal equilibrium must fulfill detailed balance, i.e.,
\begin{equation}\label{detailed_balance}
P(S_1 t_1 | S_0 t_0 ) P_{\text{eq}}(S_0) = P(S_0 t_1 | S_1 t_0 ) P_{\text{eq}}(S_1) \,.
\end{equation}
Assuming a solution of the form

\begin{equation}\label{Peq}
  P_{\text{eq}}(S)
  \propto
  e^{-\mathcal{H}(S)}\,,
\end{equation}
equation (\ref{detailed_balance}) is rewritten as

\begin{equation}\label{detailed_balance_2}
P(S_1 t_1 | S_0 t_0 ) = P(S_0 t_1 | S_1 t_0 )
\exp\left\{
        -\int_{t_0}^{t_1} dt \frac{dS}{dt}\frac{d\mathcal{H}}{dS}
    \right\}
\,,
\end{equation}
where $S(t)$ is an arbitrary path connecting $S_0$ with $S_1$.
After plugging in (\ref{pathint}) and (\ref{action}), and using the Stratonovich property, which allows us to transform a path from $S_0$ to $S_1$ into a path from $S_1$ to $S_0$ through the transformation $S(t) \rightarrow S(-t)$, we get

\begin{equation}\label{constant_rain_H}
-r_{0}+\frac{1+\gamma}{K}S
=
\frac{\gamma S^2}{2K}
\frac{d \mathcal{H}(S)}{dS}\,.
\end{equation}
The solution to (\ref{constant_rain_H}) is, up to an additive constant, given by eq.
\begin{equation}\label{solution_const_rain}
\mathcal{H}(S)=\frac{2 K r_0}{\gamma S}+\frac{2(1+\gamma)}{\gamma}\ln(S)\,.
\end{equation}
%\begin{figure}
%    \centering
%    \includegraphics[width=0.6\textwidth]{Figs/InvGamma.png}
%    \caption{Inverse gamma distributions with scale parameter $\beta = 5$ and shape parameters $\alpha = 3,4,5,6$.}
%    \label{fig:inv_gamma}
%\end{figure}
Thus, it turns out that $P_{\text{eq}}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,

\begin{equation}\label{inverse_gamma}
  P_{\text{eq}}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,,
\end{equation}
whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
\begin{equation}
  \langle S\rangle_{\eq}=Kr_{0},
\end{equation}
and whose variance, for $\gamma< 2$, is given by
\begin{equation}
  \langle (S - \langle S\rangle_{\eq})^2\rangle_{\eq}
  =
  K^2r_{0}^2
  \frac{\gamma}{2-\gamma}\,.
\end{equation}
Note that the variance diverges, for $\gamma\geq 2$.
The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.

Let us consider now the general case of a time dependent rain input $r(t)$ with the following time-series of {\em observed} outputs,
\begin{equation}\label{data}
  y_i=\frac{1}{K}S(t_i)+\sigma\epsilon_i\,,\quad i=1,\dots,n+1\,,
\end{equation}
where the times $t_i$ are chosen so that $0=t_1<t_2<\dots < t_{n+1}=T$ and the $\epsilon_i$ are uncorrelated standard normal random variables.
The {\em likelihood} function for the data (\ref{data}) is, up to a factor that doesn't depend on the model parameters, given by the path-integral
\begin{equation}\label{likelihood}
  f(\vc y | \bt)
  \propto
  \int
  \exp\bigg[
    -{\tilde{\mathcal S}}[\dot{S},S]
    -\frac{1}{2}
    \sum_{i=1}^{n+1}
    \frac
    {(y_i-S(t_i)/K)^2}
    {\sigma^2}
  \bigg]
  {\mathcal D} S(\bt)\,,
\end{equation}
with the parameter vector $\bt=(K,\gamma)^T$.
Choosing for the parameters $K$ and $\gamma$ the improper prior given by

\begin{equation}\label{prior}
  f(\bt)
  =  f(K) f(\gamma) = \frac{1}{K \gamma} \,,
\end{equation}

{\em parameter inference} means generating a sufficiently large sample of parameter vectors from the posterior
\begin{equation}\label{posterior}
  f(\bt|\vc y)\propto f(\vc y|\bt)f(\bt)\,.
\end{equation}

We replace the variable $S(t)$ and the parameter $K$ by the dimensionless quantities, $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
In terms of these dimensionless variables and parameters the posterior reads (without prior)
\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -{{\mathcal S}}[\dot{q},q]
    -\frac{1}{2}
    \sum_{i=1}^{n+1}
    \frac
    {(y_i-r_ie^{\beta q_i})^2}
    {\sigma^2}
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,,
\end{equation}
%\prod_{t}
%\frac{dS(t)}{\sqrt{2\pi \mathit{\Delta t} \gamma/K }S(t)}\,,
where $r_i$ and $q_i$ represent $r(t)$ and $q(t)$, respectively, at time $t_i$. In the transformed variables the path-measure takes the form, up to a parameter-independent multiplicative factor,

\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,,
\end{equation}

while the action is rewritten in the compact form,

\begin{equation}\label{action_new}
{\mathcal S}[\dot{q},q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\} \,,
\end{equation}

with
\begin{equation}\label{rho}
\rho(t)=\frac{T\dot r(t)}{\beta r(t)}+\frac{\beta(\gamma+1)}{\gamma}
=
\frac{T}{\beta}\frac{d \left[\ln r(t)\right]}{dt}+\frac{\beta(\gamma+1)}{\gamma}\,.
\end{equation}

In order to ease the numerical computations, it is useful to introduce the time-dependent Hamiltonian,

\begin{equation}\label{H}
  \mathcal{H}_q(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}

and rewrite the action (\ref{action_new}) as,

\begin{multline}\label{action_final}
{\mathcal S}[\dot{q},q]
=
\frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}_q(q,t)}{\partial t} -
     \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+ \mathcal{H}_q(q(T),T) - \mathcal{H}_q(q(0),0)
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{multline}

Since the integral on the r.h.s. of (\ref{posterior_pathint}) is very expensive to calculate, we will sample simultaneously parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
Since this distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by \cite{duane_1987_HMC}.
This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system. Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian

\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}

where

\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m_q}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}

where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The HMC method, which is a combination of the {\em Metropolis algorithm} by \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps

\begin{enumerate}
  \item
  Momenta are sampled from the Gaussian distribution defined by (\ref{K}).
  \item
  A volume-preserving and time-reversible solution to a discretization of Hamilton's equations is calculated numerically (leap-frog algorithm).
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}

The last step is the standard Metropolis algorithm, while the first two steps allow to make relatively large jumps in phase space while maintaining a relatively large acceptance rate.

Let us first write down the proper discretization of the path-integral (\ref{posterior_pathint}), with the action (\ref{action_final}).
Therefore, let us assume that the measurement time points $\left\{ y_i \right\}_{i=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
The discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$ read

\begin{align}
   K( p,{\pmb\pi})
   &\approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_q}\Delta t
   +
   \sum_{\alpha=1}^s\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt)
   &\approx
   \frac{\Delta t}{T}
   \sum_{i=2}^{N}
   \left\{
    \frac{1}{2} T^2 \dot q_i^2
    +
   \left(
        \frac{1}{2}
        \left(
            \rho_i-\frac{\beta}{\gamma}e^{-\beta \bar q_i}
        \right)^2
        -
        T\bar q_i\dot\rho_i
    \right)
   \right\}
  \\
  &-
  \frac{(2+\gamma)\beta^2}{4\gamma}
  +
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N-1}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  & +
  \sum_{s=1}^{n+1}
  \frac{(y_s-r_{(s-1)j+1} e^{\beta q_{(s-1)j+1}})^2}{2\sigma^2}
  % +\ln \left(K^{1-N/2} \gamma^{1+N/2}\right)\nonumber
   \,,\label{Vdisc}
\end{align}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,\quad
  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = K \frac{\ln(r_{i+1}/r_i)}{\Delta t}+\gamma+1\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}

Note that we've neglected terms of order $\mathcal O(1/N)$ in the action (\ref{Vdisc}) and that we've assumed $r(t)$ to be at least twice differentiable.

@@

We choose another approach, the so-called \emph{staging} as described in Tuckerman et al. (1993), and diagonalize the harmonic part only in between the measurement points.
We first note that the Hamiltonian $\mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})$ (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$. 
Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom. 
Sampling system realizations, each defined by a set of coordinates $\left\{ q_i\right\}$ and parameters $\bt$, is equivalent to sampling states in the polymer dynamics.

The discretized harmonic part of the action is rewritten as

\begin{multline}
  \sum_{i=2}^{N}
  \frac{T}{2\Delta t}
  (q_i-q_{i-1})^2
  \\ =
  \frac{T}{2}
  \sum_{s=1}^{n}\left\{
    \frac{(q_{(s-1)j+1} - q_{sj+1})^2}{j\Delta t}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)\Delta t}
    (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
  \right\}\,,
\end{multline}

with

\begin{equation}
  q^*_{(s-1)j+k}
  =
  \frac{(k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} }{k}
  \,.
\end{equation}

Following Tuckerman et al. (1993) we apply the coordinate transformation

\begin{equation}
  u_{sj+1} = q_{sj+1}\,,\quad
  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,,
\end{equation}

with its inverse given by

\begin{equation}
  q_{sj+1} = u_{sj+1}\,,\quad
  q_{sj+k} = \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
  +\frac{j-k+1}{j}u_{sj+1}\,,
\end{equation}

or equivalently by the recursive relation

\begin{equation}
  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{s+k+1}+ \frac{1}{k}q_{sj+1} \,.
\end{equation}

>>>>

In this paper, we study the feasibility of a full Bayesian inference, for simple conceptual rainfall-runoff models with scale-invariant noise.
We start with a single linear reservoir and describe the dynamics of its water content, $S(t)$, by means of the stochastic differential equation
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ describes the rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Due to the state-dependence of the noise term, eq. (\ref{sde}) is ill-defined unless we specify the discretization convention.
For computational reasons, we work with the {\em Stratonovich} convention.
Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{eq}=Kr_0$ (see below).

The scale-invariance of the noise term renders the noise parameter $\gamma$ dimensionless.
Before we proceed, we replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
W.r.t. these new  variables and parameters, model equation (\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with
\begin{equation}
  \rho(t)
  =
  \frac{T}{\beta}\frac{d}{dt}\ln r(t)
  +
  \frac{(2+\gamma)\beta}{2\gamma}\,.
\end{equation}

%Given rainfall and runoff time-series, we want to infer the two system parameters $K$ and $\gamma$, measuring, respectively, the retention time and the noise.

According to \cite{lau_2007_StateDepDiff}, the probability, $P(q_1,T|q_0,0)$, of finding the system in a state $q$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, can be expressed in the form of a path integral as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0) 
= 
\frac{1}{Z}
\int 
e^{-{\mathcal S}[q,\dot q]} 
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
The path-measure $\mathcal Dq$ is formally written as an infinite product
\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,.
\end{equation}
The {\em action} is a functional on the space of paths and reads as
%\begin{equation}\label{action}
%  \tilde{\mathcal S}[\dot{S}(t),S(t)]
%  =
%  \int_0^T
%  dt
%  \left\{
%    \frac{K}{2\gamma S^2(t)}
%    \left(
%        \dot S(t)
%        -
%        r(t)
%        +
%        \frac{1+\gamma}{K}S(t)
%    \right)^2
%    -
%    \frac{2+\gamma}{4K}
%  \right\}
% \,.
%\end{equation}
\begin{equation}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.

We introduce the time-dependent {\em Hamiltonian}
\begin{equation}\label{H}
  \mathcal{H}(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}
and rewrite action (\ref{action}) as,
\begin{multline}\label{action}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0)
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{multline}

Next, we calculate the equilibrium distribution $P_{\text{eq}}(q) = \lim_{T\rightarrow\infty} P(q,T|q_0,0)$ in the simple case of a constant rain input $r(t)\equiv r_{0}$. Systems at thermal equilibrium must fulfill detailed balance, i.e.,
\begin{equation}\label{detailed_balance}
P(q_1 t_1 | q_0 t_0 ) P_{\text{eq}}(q_0) = P(q_0 t_1 | q_1 t_0 ) P_{\text{eq}}(q_1) \,.
\end{equation}
After plugging in (\ref{pathint}) and (\ref{action}), and using the transformation $q(t) \rightarrow q(-t)$, which maps paths from $q_0$ to $q_1$ to paths from $q_1$ to $q_0$, we get, since $\dot\rho(t)= 0$,
\begin{equation}\label{Peq}
  P_{\text{eq}}(q)
  \propto
  e^{-2\mathcal{H}(q)}\,.
\end{equation}
Transforming back to the original variables, it turns out that $P_{\text{eq}}(S)$ is described by an inverse gamma distribution with scale parameter $2Kr_{0}/\gamma$ and shape parameter $(2+\gamma)/\gamma$, i.e.,
\begin{equation}\label{inverse_gamma}
  P_{\text{eq}}(S)
  \propto
  S^{-2(1+\gamma)/\gamma}e^{-2Kr_{0}/(\gamma S)}\,,
\end{equation}
whose mean equals the equilibrium solution of the unperturbed system ($\gamma=0$),
\begin{equation}
  \langle S\rangle_{\eq}=Kr_{0},
\end{equation}
and whose variance, for $\gamma< 2$, is given by
\begin{equation}
  \langle (S - \langle S\rangle_{\eq})^2\rangle_{\eq}
  =
  K^2r_{0}^2
  \frac{\gamma}{2-\gamma}\,.
\end{equation}
Note that the variance diverges, for $\gamma\geq 2$.
The power-law decay of the inverse gamma distribution is reminiscent of the scale-invariance of the error model.

Let us now consider two time-series of {\em observed} rain inputs and discharge outputs, $r_s=r(t_s)$ and $y_s=S(t_s)/K$, respectively, measured at times $0=t_1<t_2<\dots < t_{n+1}=T$.
For the measurement error model, we simply set
\begin{equation}\label{data}
  \ln\frac{y_s}{r_s}
  =
  \beta q_s
  +
  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,,
\end{equation}
where $\epsilon_s$ are uncorrelated standard normal random variables.

%The {\em likelihood} function for the data (\ref{data}) is, up to a factor that doesn't depend on the model parameters, given by the path-integral
%\begin{equation}\label{likelihood}
%  f(\vc y | \bt)
%  \propto
%  \int
%  \exp\bigg[
%    -{\tilde{\mathcal S}}[\dot{S},S]
%    -\frac{1}{2}
%    \sum_{i=1}^{n+1}
%    \frac
%    {(y_i-S(t_i)/K)^2}
%    {\sigma^2}
%  \bigg]
%  {\mathcal D} S(\bt)\,,
%\end{equation}
%with the parameter vector $\bt=(K,\gamma)^T$.
%Choosing for the parameters $K$ and $\gamma$ the improper prior given by
%
%\begin{equation}\label{prior}
%  f(\bt)
%  =  f(K) f(\gamma) = \frac{1}{K \gamma} \,,
%\end{equation}
%
%{\em parameter inference} means generating a sufficiently large sample of parameter vectors from the posterior
%\begin{equation}\label{posterior}
%  f(\bt|\vc y)\propto f(\vc y|\bt)f(\bt)\,.
%\end{equation}

The goal is to infer the posterior parameter distribution, for the parameters $\bt=(\beta,\gamma)^T$, based on the measured time series $\vc r$ and $\vc y$.
For the time being we assume constant priors.
Then, the posterior reads as
\begin{equation}\label{posterior_pathint}
  f(\bt | \vc y, \vc r)
  \propto
  \int
  \exp\bigg[
    -{{\mathcal S}}[{q},\dot q]
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r_s)-{\beta q_s})^2}
    {\sigma^2}
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
Since the integral on the r.h.s. of (\ref{posterior_pathint}) is very expensive to calculate, we will sample, simultaneously, parameter vectors $\bt$ and discretized system realizations $q(t)$ directly from an appropriate discretization of the kernel of (\ref{posterior_pathint}).
Since the associated distribution is very high dimensional and strongly correlated, we employ the method called {\em Hamiltonian Monte Carlo} (HMC) introduced by \cite{duane_1987_HMC}.
This method interprets the exponent of the kernel of the posterior as the potential of a one-dimensional statistical mechanical system. Each degree of freedom, $q(t)$ and $\bt$ in our case, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$, respectively, and the system is defined by the  Hamiltonian
\begin{equation}\label{Hamiltonian}
    \mathcal{H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})
    =
    K( p,{\pmb\pi}) + V( q,\bt)\,,
\end{equation}
with
\begin{equation}\label{K}
   K( p,{\pmb\pi})
   =
   \int_0^T \frac{ p^2(t)}{2m_q}dt
   + \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,
\end{equation}
where $V( q,\bt)$ is the negative logarithm of the kernel of (\ref{posterior_pathint}).
The posterior (\ref{posterior_pathint}) is expressed by the phase space path integral
\begin{equation}\label{phaseSpacePathInt}
    f(\bt | \vc y, \vc r)
  \propto
  \int
  e^{-\mathcal {H}_{\text{HMC}}(q,\bt; p,{\pmb\pi})}
  {\mathcal Dp}
  \wedge
  {\mathcal Dq}
  \wedge
  d{\pmb\pi}
  \,.
\end{equation}

The HMC method, which is a combination of the {\em Metropolis algorithm} by \cite{metropolis_1953_MRT2} and {\em molecular dynamics} methods \textbf{[ref?]}, iterates the following steps
\begin{enumerate}
  \item
  Momenta are sampled from the Gaussian distribution defined by (\ref{K}).
  \item
  A volume-preserving and time-reversible solution to a discretization of Hamilton's equations is calculated numerically.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make relatively large jumps in phase space while maintaining a relatively large acceptance rate.

Let us first write down the discretization of the path-integral (\ref{phaseSpacePathInt}).
Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
Path-integral (\ref{posterior_pathint}) is approximated by an ordinary integral, with the approximate path-measure 
\begin{equation}
  \mathcal Dp\wedge\mathcal Dq
  \approx
  \prod_i dp_i\wedge dq_i 
\end{equation}
and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
\begin{align}
   K( p,{\pmb\pi})
   &\approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_q}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt)
   &\approx
   \frac{\Delta t}{T}
   \sum_{i=2}^{N}
   \left\{
    \frac{1}{2} 
    T^2 \dot q_i^2
    +
    \frac{1}{2}
     \left(
        \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i}
     \right)^2
    -
    \frac{\beta^2}{2\gamma}
    e^{-\beta q_i}
   -
    T q_i\dot\rho_i
   \right\}
  \\
  &+
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  & +
  \sum_{s=1}^{n+1}
  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
   \,,\label{Vdisc}
\end{align}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
  %\quad
%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = \frac{T}{\beta} \frac{\ln(r_{i+1}/r_i)}{\Delta t}
+
\frac{(2+\gamma)\beta}{2\gamma}
\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}
Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}) and that we've assumed $r_i$ to be the discretization of an at least twice differentiable interpolation of the input measurements.

@@

We choose another approach, the so-called \emph{staging} as described in \cite{tuckerman1992reversible}, and diagonalize the harmonic part only in between the measurement points.
We first note that Hamiltonian (\ref{Hamiltonian}) can be seen as the classical Hamiltonian of a polymer chain with harmonic bonds between neighbouring beads, described by the first harmonic term of $V(q,\bt)$, in an external field described by the other non-harmonic terms of $V(q,\bt)$.
Furthermore, the model parameters are viewed as additional degrees of freedom that couple with all other degrees of freedom.
Sampling system realizations, each defined by a set of coordinates $\left\{ q_i\right\}$ and parameters $\bt$, is equivalent to sampling states in the polymer dynamics.

The discretized harmonic part of the action is rewritten as
\begin{multline}
  \sum_{i=2}^{N}
  \frac{T}{2\Delta t}
  (q_i-q_{i-1})^2
  \\ =
  \frac{T}{2}
  \sum_{s=1}^{n}\left\{
    \frac{(q_{(s-1)j+1} - q_{sj+1})^2}{j\Delta t}
    +
    \sum_{k=2}^j
    \frac{k}{(k-1)\Delta t}
    (q_{(s-1)j+k}-q^*_{(s-1)j+k})^2
  \right\}\,,
\end{multline}
with
\begin{equation}
  q^*_{(s-1)j+k}
  =
  \frac{(k-1)q_{(s-1)j+k+1} + q_{(s-1)j+1} }{k}
  \,.
\end{equation}
Following \cite{tuckerman1993efficient} we apply the coordinate transformation
\begin{equation}
  u_{sj+1} = q_{sj+1}\,,\quad
  u_{sj+k} = q_{sj+k} - q^*_{sj+k}\,,
\end{equation}
with its inverse given by
\begin{equation}
  q_{sj+1} = u_{sj+1}\,,\quad
  q_{sj+k} = \sum_{l=k}^{j+1}\frac{k-1}{l-1}u_{sj+l}
  +\frac{j-k+1}{j}u_{sj+1}\,,
\end{equation}
or equivalently by the recursive relation
\begin{equation}
  q_{sj+k} = u_{sj+k} + \frac{k-1}{k} q_{s+k+1}+ \frac{1}{k}q_{sj+1} \,.
\end{equation}
%Moreover, we introduce the following transformation for the parameters,
%
%\begin{equation}\label{params_transform}
%\beta = \ln(K/T) \,,\quad \tau=\ln(\gamma)
%\end{equation}
%
%and the corresponding transformed vector $\tilde{\bt}=(\beta,\tau)^T$. This transformation ensures that the system parameters $K$ and $\gamma$ will be kept positive during the HMC algorithm.
%
